Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
Tap

Our main metatheoretic results will be \textit{substitution} and
\textit{label-substitution}, which will require introducing some new
metatheoretic judgements to state properly, which we do in Figure
\ref{fig:ssa-meta-judgements}.

@@

In particular, the judgement \(\issubst{\gamma}{\Theta}{\Gamma}\) says that
\(\gamma\), which is taken to be a map from variable names to expressions, is a
valid substitution taking \(\Theta\) to \(\Gamma\). What this means is that, for
each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted
as a \textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such
that when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine
variables in \(\Theta\) are used up: this is described formally via the typing
rules \rle{subst-nil}, which says that any function \(\gamma\) takes the empty
context to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).

@@

We can now state our substitution theorem for terms in a straightforward way: if
\(\substctx{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
\(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and
\(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{e}}{A}{q}\).
Unfortunately, the substitution theorem for blocks is a little more complicated:
the naive choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may contain
captured variables from \(\Gamma\) that are no longer available if we are typing
\([\gamma]t\) in the context of \(\Theta\). To solve this problem, we need to
introduce the concept of a \textit{label-substitution} \(\mc{L}\) \todo{rest}

>>>>

@@

Our main metatheoretic results will be \textit{substitution} and
\textit{label-substitution}, which will require introducing some new
metatheoretic judgements to state properly, which we do in Figure
\ref{fig:ssa-meta-judgements}. In particular, the judgement
\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is taken to be
a map from variable names to expressions, is a valid substitution taking
\(\Theta\) to \(\Gamma\). What this means is that, for each variable
\(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted as a
\textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such that
when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
in \(\Theta\) are used up: this is described formally via the typing rules
\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).

@@

We can now state our substitution theorem for terms in a straightforward way: if
\(\issubst{\gamma}{\Theta}{\Gamma}\) and \(\hasty{\Gamma}{p}{e}{A}{q}\), then
\(\hasty{\Theta}{p}{[\gamma]e}{A}{q}\) and
\(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{e}}{A}{q}\).
Unfortunately, the substitution theorem for blocks is a little more complicated:
the naive choice of attempting to show \(\haslb{\Gamma}{t}{\ms{L}} \implies
\haslb{\Theta}{[\gamma]t}{\ms{L}}\) does not work, since \(\ms{L}\) may contain
captured variables from \(\Gamma\) that are no longer available if we are typing
\([\gamma]t\) in the context of \(\Theta\). To solve this problem, we need to
introduce the concept of a \textit{label-substitution} \(\mc{L}\) \todo{rest}
